Reassessing the cost-effectiveness of pneumococcal conjugate vaccines in low- and middle-income countries using opportunity cost-based thresholds

### 2024-07-22 manuscript.tex
Updates from Overleaf
Updated

One component of HTA process is cost-effectiveness analysis (CEA). CEA is method for weighing costs and outcomes in comparable method across interventions and thus provides guidance on value for money of new technologies. Results are typically presented as incremental cost-effectiveness ratios, in \$/QALY gained or \$/DALY averted. If a vaccine is found to be cost-saving, usually implemented pending other factors (budget impact, political and equity considerations, and other practical and logistical considerations including availability of vaccine, etc). Vaccines that are not cost-saving, but result in higher costs and higher health benefits, are often compared to country-specific willingness to pay thresholds, to determine whether an intervention is considered cost-effective (that is, of good value for money). Recent systematic reviews PCV is CE across (most)( LMICs (sometimes cost-saving) \supercite{saokaew_cost_2016, syeed_pneumococcal_2023, wang_systematic_2022, zakiyah_pneumococcal_2020}. Cost-effectiveness is influenced by vaccine price (local acquisition), inclusion/exclusion of indirect effects (serotype replacement and herd effects), cross protection, disease burden (Incidence / mortality from invasive pneumococcal disease (IPD), and pneumonia), and vaccine efficacy and coverage

Historically thresholds of 1-3 x gross domestic product (GDP) per capita have been used to assess the cost-effectiveness of vaccines and other interventions in low- and middle-income countries. They were proposed by the World Health Organization (WHO) Commission on Macroeconomics and Health (CMH) based on human capital considerations rather than health opportunity costs [REF]. Hence, they represented what the CMH believed governments should pay for improved health, rather than what interventions to choose given a fixed healthcare budget. They became widely cited and used to evaluate individual interventions, but WHO produced a paper advising against the use of GDP-based thresholds, especially in isolation from other considerations \parencite{bertram_costeffectiveness_2016}. Despite the World Health Organization and others recommending against using GDP per capita-based thresholds for judging the value for money of healthcare interventions most recent studies continue to do so4. Recent systematic review of 230 studies (713 interventions) published CEAs in LMICs: 1-3 × GDP/ capita was the most common type of threshold used (84.3\%) ; ~ A third of studies (34.2\%) using 1 to 3× GDP per capita applied a threshold at 3× GDP per capita. No study used locally developed thresholds. 79.3\% of interventions received a recommendation as “cost-effective”.

Newer country-specific estimates of cost-effectiveness thresholds (CETs) exist based on health opportunity costs are available. \supercite{ochalek_estimating_2018, pichon-riviere_determining_2023,woods_country-level_2016} New approach to estimate threshold using empirical estimates of investments (supply side) or WTP (demand side). Several countries are in process of developing these thresholds (ex: Indonesia ). New estimates from \textcite{woods_country-level_2016} (based on one empirical estimate from a high-income setting), \textcite{ochalek_estimating_2018} (using multiple empirical data points, from LMICs), and \textcite{pichon-riviere_determining_2023} provide updated guidance on health opportunity costs in LMICs. \textcite{woods_country-level_2016} also provides these estimates for high-income countries. The main aim of this study is to investigate whether these estimates on cost-effectiveness thresholds (CET) based on health opportunity costs may impact on study conclusions about the cost-effectiveness of pneumococcal conjugate vaccine programs in low- and middle-income countries. A secondary aim is to investigate which study factors are influential in driving study conclusions that switch from finding PCV to be cost-effective to not cost-effective.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Methods}
\subsection{Systematic search}
We undertook a systematic literature search using similar search strings as a previous review on pneumococcal vaccination in children.\supercite{saokaew_cost_2016} We used the search strings “pneumococc* AND conjugat* AND (vaccin* OR immun*) AND “economic OR cost-effectiveness OR cost-benefit OR cost-utility OR cost-effectiveness OR cost-benefit OR cost-utility” in the title, abstract, and keyword fields of Web of Science, Scopus, PubMed, and EmBase (Ovid) from inception until 25 April 2022. The search was limited to low- and middle-income countries (LMICs) using country names and related LMIC terminology (see Appendix 1 in Supplementary Material). We also checked published systematic reviews on economic evaluation of PCV for additional studies.\supercite{saokaew_cost_2016, syeed_pneumococcal_2023, wang_systematic_2022, zakiyah_pneumococcal_2020}

All identified titles and abstracts were screened by one author (EA) to remove studies that were clearly not related to the topic, were not primary research, or not focused on a LMIC. Two reviewers (EA, NC) independently screened the full text of studies remaining against seven inclusion criteria: (1) original articles containing a full economic evaluation (two evaluations, costs and consequences ),5 (2) focused on one or more LMICs, (3) evaluated PCV covering seven or more serotypes, (4) evaluated PCV in children less than 12 years, (5) used either disability-adjusted life-years (DALYs), quality-adjusted life years (QALYs) or life-years (LY), as outcome measure, (6) reported an incremental cost-effectiveness ratio (ICER) and (7) were in English. Editorials, letters, conference or meeting abstracts or proceedings, and review articles were excluded. Multi-country studies not reporting country-specific ICERs were also excluded. Evaluations of pneumococcal polysaccharide vaccine were excluded.

The following data were extracted by one reviewer (EA) into an Excel data extraction template: study characteristics (country, currency and cost year, World Bank country income group, vaccine evaluated and comparator, number and unit price of doses); methodological assumptions of cost-effectiveness analysis (perspective (societal or healthcare payer), time horizon, discount rate, and considerations of herd effects (with or without) and serotype replacement (with or without)); and study results (outcome measure, ICER for each evaluation where this was separately reported (by perspective, and by inclusion or exclusion of herd effects and by serotype replacement, where possible), inclusion of budget impact (yes or no), cost-effectiveness threshold used and conclusion on cost-effectiveness, for each evaluation ). A second reviewer (NC) double-extracted one-third of studies. Any inconsistencies were discussed resolved in consultation with both reviewers. The extracted data was compared with data from published systematic reviews1-4 to verify the extracted data on study characteristics, parameters and ICERs.

Risk of bias was not assessed as this has been done by previously published systematic reviews2 and was not the focus of the current review.

\subsection{Data preparation}
Countries were classified by World Bank country income group as upper middle (UM), lower middle (LM), and low (L) income. Where reported in a different currency, ICERs and vaccine dose prices were converted to United States dollars (USD) using year- and currency-specific exchange rates from the World Bank.6 Country- and year-specific GDP per capita estimates (in USD) were compiled from World Bank data and used to convert ICERs for all evaluations into a \% GDP per capita. We extracted country-specific upper- and lower-bound CETs from the Woods7 and Ochalek8 studies, in terms of \% GDP per capita, using the highest and lowest estimates from both studies. We then generated updated study conclusions (“cost-saving”, “cost-effective” or “not-cost-effective) for each evaluation according to the Woods and Ochalek CETs, by comparing the converted ICER (in \% GDP per capita) to the upper- and lower-bound CETs (in \% GDP per capita ). We generated four indicator variables to indicate whether study conclusions shifted from being “cost-effective” based on study specific threshold and conclusion, to “not cost-effective” (based on each of the four updated CET estimates: WoodsHigh, WoodsLow, OchalekHigh, OchalekLow). Studies where CETs were not explicitly stated were not included in the statistical analysis.

\subsection{Statistical analysis}
We first summarised the key characteristics of studies and unique evaluations included in the review including distribution of CETs used and study conclusions. We then conducted a descriptive analysis of the evaluations that switched from being considered “cost-effective” by the study conclusion to “not cost-effective” (across any of the 4 CETs), by study perspective, type of comparison, income group, considerations of herd effects or serotype replacement, vaccine dose price by income group.

Next, we conducted two-sample tests of differences in vaccine dose price, country income group, perspective of analysis, and considerations of herd immunity  or serotype replacement among evaluations that switched from being cost-effective across any of the 4 CETs, compared to those that didn’t switch.

@@

\clearpage
\newrefcontext[sorting=nyt]
\printbibliography

>>>>

One component of HTA process is cost-effectiveness analysis (CEA). CEA is method for weighing costs and outcomes in comparable method across interventions and thus provides guidance on value for money of new technologies. Results are typically presented as incremental cost-effectiveness ratios, in \$/QALY gained or \$/DALY averted. If a vaccine is found to be cost-saving, usually implemented pending other factors (budget impact, political and equity considerations, and other practical and logistical considerations including availability of vaccine, etc). Vaccines that are not cost-saving, but result in higher costs and higher health benefits, are often compared to country-specific willingness to pay thresholds, to determine whether an intervention is considered cost-effective (that is, of good value for money). Recent systematic reviews PCV is CE across (most)( LMICs (sometimes cost-saving) \autocite{saokaew_cost_2016, syeed_pneumococcal_2023, wang_systematic_2022, zakiyah_pneumococcal_2020}. Cost-effectiveness is influenced by vaccine price (local acquisition), inclusion/exclusion of indirect effects (serotype replacement and herd effects), cross protection, disease burden (Incidence / mortality from invasive pneumococcal disease (IPD), and pneumonia), and vaccine efficacy and coverage

Historically thresholds of 1-3 x gross domestic product (GDP) per capita have been used to assess the cost-effectiveness of vaccines and other interventions in low- and middle-income countries. They were proposed by the World Health Organization (WHO) Commission on Macroeconomics and Health (CMH) based on human capital considerations rather than health opportunity costs [REF]. Hence, they represented what the CMH believed governments should pay for improved health, rather than what interventions to choose given a fixed healthcare budget. They became widely cited and used to evaluate individual interventions, but WHO produced a paper advising against the use of GDP-based thresholds, especially in isolation from other considerations \autocite{bertram_costeffectiveness_2016}. Despite the World Health Organization and others recommending against using GDP per capita-based thresholds for judging the value for money of healthcare interventions most recent studies continue to do so4. Recent systematic review of 230 studies (713 interventions) published CEAs in LMICs: 1-3 × GDP/ capita was the most common type of threshold used (84.3\%) ; ~ A third of studies (34.2\%) using 1 to 3× GDP per capita applied a threshold at 3× GDP per capita. No study used locally developed thresholds. 79.3\% of interventions received a recommendation as “cost-effective”.

Newer country-specific estimates of cost-effectiveness thresholds (CETs) exist based on health opportunity costs are available. \autocite{woods_country-level_2016, ochalek_estimating_2018, pichon-riviere_determining_2023} New approach to estimate threshold using empirical estimates of investments (supply side) or WTP (demand side). Several countries are in process of developing these thresholds (ex: Indonesia ). New estimates from \autocite{woods_country-level_2016} (based on one empirical estimate from a high-income setting), \autocite{ochalek_estimating_2018} (using multiple empirical data points, from LMICs), and \autocite{pichon-riviere_determining_2023} provide updated guidance on health opportunity costs in LMICs. \autocite{woods_country-level_2016} also provides these estimates for high-income countries. The main aim of this study is to investigate whether these estimates on cost-effectiveness thresholds (CET) based on health opportunity costs may impact on study conclusions about the cost-effectiveness of pneumococcal conjugate vaccine programs in low- and middle-income countries. A secondary aim is to investigate which study factors are influential in driving study conclusions that switch from finding PCV to be cost-effective to not cost-effective.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Methods}
\subsection{Literature search strategy}
We undertook a systematic literature search using similar search strings as a previous review on pneumococcal vaccination in children.\autocite{saokaew_cost_2016} We used the search strings “pneumococc* AND conjugat* AND (vaccin* OR immun*) AND “economic OR cost-effectiveness OR cost-utility” in the title, abstract, and keyword fields of Web of Science, Scopus, PubMed, and EmBase (Ovid) from inception until 17 July 2024. The search was limited to low- and middle-income countries (LMICs) using country names and related LMIC terminology (see Appendix 1 in Supplementary Material). We also checked published systematic reviews on economic evaluation of PCV for additional studies.\autocite{saokaew_cost_2016, syeed_pneumococcal_2023, wang_systematic_2022, zakiyah_pneumococcal_2020}

\subsection{Inclusion and exclusion strategy}
All identified titles and abstracts were screened by one author (EA) to remove studies that were clearly not related to the topic, were not primary research, or not focused on a LMIC. Two reviewers (EA and NC) independently screened the full text of studies remaining against seven inclusion criteria: (i) original articles containing a cost-effectiveness or cost-utility analysis, (ii) focused on one or more LMICs, (iii) evaluated PCV covering seven or more serotypes, (iv) evaluated PCV in children less than 12 years, (v) used either disability-adjusted life-years (DALYs), quality-adjusted life years (QALYs) or life-years (LY), as outcome measure, (vi) reported an incremental cost-effectiveness ratio (ICER) and (vii) were written in English. Editorials, letters, conference or meeting abstracts or proceedings, and review articles were excluded. Multi-country studies not reporting country-specific ICERs were also excluded. Evaluations of pneumococcal polysaccharide vaccine were excluded.

\subsection{Data extraction}
The characteristics, methodological decisions, and results from each study were extracted by one reviewer (EA) into an Excel workbook. Study characteristics included the country, currency and cost year, World Bank country income group, type of vaccine evaluated (PCV7, PCV9, PCV10, or PCV13), and the number of doses. Countries were classified by World Bank country income group as upper middle (UM), lower middle (LM), and low (L) income. We then extracted information on the following parameters and methodological assumptions: study perspective (societal or healthcare/payer), time horizon, discount rate, and inclusion/exclusion of herd effects, inclusion/exclusion of serotype replacement, and unit price per vaccine used. We categorized the study results into cost-saving, cost-effective, or not-cost-effective. Where reported in a different currency, vaccine dose prices and ICERs were converted to United States dollars (USD) using year- and currency-specific exchange rates from the World Bank.6

Most studies reported ICERs for more than one vaccine comparison or combination of methodological assumptions. For example, a study could report results for both PCV7 and PCV10 compared to no vaccination (NoVax), with and without including herd effects in both comparisons. This implies for that single study, there will be four evaluations (PCV7 vs NoVax with herd effects, PCV10 vs NoVax with herd effects, PCV7 vs NoVax without herd effects, and PCV10 vs NoVax without herd effects). Where this was the case, we extracted these ICERS and study characteristics in separate Excel rows. The extracted data was compared with data from published systematic reviews\autocite{saokaew_cost_2016, syeed_pneumococcal_2023, wang_systematic_2022, zakiyah_pneumococcal_2020} to verify the information on study characteristics, parameters, and ICERs. A second reviewer (NC) double-extracted one-third of studies. Any inconsistencies were discussed resolved in consultation with both reviewers. Risk of bias was not assessed as this has been done by previously published systematic reviews\textbf{2} and was not the focus of the current review.

% see https://www.tandfonline.com/doi/full/10.1080/14760584.2023.2173176
\subsection{Cost-effectiveness threshold calculations}
We updated the cost-effectiveness results for each evaluation from the studies as follows. First, we calculated the ICER as a \% of GDP per capita by dividing each ICER by the country- and year-specific GDP per capita from The World Bank \autocite{the_world_bank_gdp_2023}. Next, we extracted country-specific highest and lowest CET estimates from the three studies \autocite{woods_country-level_2016, ochalek_estimating_2018, pichon-riviere_determining_2023}. We then converted these CETs to equivalent \% GDP per capita, using the year-specific GDP per capita \autocite{the_world_bank_gdp_2023}: 2013 for Woods\autocite{woods_country-level_2016} and 2019 for P-Riverie\autocite{pichon-riviere_determining_2023}. The CET as \% GDP per capita was obtained directly from Ochalek\autocite{ochalek_estimating_2018}. Finally, we then generated updated study conclusions (“cost-effective” or “not-cost-effective) for each evaluation according to the Woods and Ochalek CETs, by comparing the converted ICER (in \% GDP per capita) to the upper- and lower-bound CETs (in \% GDP per capita ). The results of cost-saving evaluations remained unchanged as they do not depend on the CET value.

 We generated six indicator variables to record whether study conclusions shifted from being “cost-effective” based on study-specific threshold and conclusion, to “not cost-effective”. These variables were based on each of the updated CET estimates: Woods (2016) High value, Woods (2016) Low value, Ochalek (2018) High value, Ochalek (2018) Low value, P-Rivere (2023) High value, and P-Rivere (2023) High value. We also generated an "any switch" variable, to indicate switching from being cost-effective to not-cost-effective across any of the six CET estimatess. to for  Studies where CETs were not explicitly stated were excluded from the statistical analysis.

\subsection{Statistical analysis}
We first summarised the key characteristics of studies and unique evaluations included in the review including distribution of CETs used and study conclusions. We then conducted a descriptive analysis of the evaluations that switched from being considered “cost-effective” by the study conclusion to “not cost-effective” (across any of the six CETs), by study perspective, type of comparison, income group, considerations of herd effects or serotype replacement, vaccine dose price by income group. Next, we conducted statistical tests for differences among evaluations that switched from being cost-effective across any of the six CETs, compared to those that didn’t switch. The tests used were two-sample tests for differences in vaccine dose price, and Fisher's exact test for differences by country income group, perspective of analysis, and inclusion of herd immunity or serotype replacement.

@@

\clearpage
\printbibliography
